---
title: Iterating to a Better Visualization
author: Ryan Thomas
date: '2019-02-17'
slug: iterating-to-a-better-visualization
categories: []
tags: []
---


A few months ago I was presented with a visualization at work by a colleague 
who was very excited about it. They were adamant that I add it to the analytics 
package I was working on. I thought there several problems with the 
visualization. In this post, I have recount the iterations I went through to
create what I think is a better visulation. 


## Where It All Began

All of the visualizations are trying to communicate the tradeoffs 
between two alternative set of projected cashflows. One, the baseline, is the 
set of cashflows which will occur if the client continues with their current 
policy unchanged. The alternative is thet set of cashflows which is projected to
occur if the client implements a new policy. 

In the d_raw table, the delta column is the difference in the aggregate 
annual cashflows for alternative policy and the baseline policy. I'm starting 
with the differences because that was the only data I had in the original 
visualization. The data below is not the same as the original but is 
analogous. 

```{r}
suppressPackageStartupMessages({
        library(tidyverse)
        library(glue)
        library(gt)
        })

d_raw <- tribble(
  ~py, ~delta,
  2018,  -248,
  2019,  -193,
  2020,  -121,
  2021,   -53,
  2022,   -19,
  2023,    75,
  2024,   222,
  2025,   310,
  2026,   425
)

d <- mutate(d_raw, 
            scenario = as.factor(case_when(delta > 0 ~ "alternative", 
                                           TRUE ~ "baseline")),
            label = glue("${abs(delta)}")
)
gt(d)
```

```{r}

ggplot(d, aes(x = py)) + 
        geom_point(aes(y = 1, size = abs(delta), color = scenario)) +
        scale_size(range = c(5, 22)) +
        geom_text(aes(y = 1, label = label)) + 
        scale_color_manual(values = c("#8EE5EE", "#AB82FF")) +
        geom_vline(xintercept = c(2022 + 2/3, 2025 + 2/3), 
                   linetype = "dashed") +
        geom_label(aes(x = 2022, y = 1.45, label = "Breakeven 68\nmonths")) +
        geom_label(aes(x = 2024.9, y = 1.45, 
                       label = "Realized gains\nbegin month 104")) +
        geom_segment(aes(x = 2017.5, xend = 2022.5, y = 1.2, yend = 1.2), 
                     color = "blue") +
        geom_text(aes(x = 2020, y = 1.27), 
                  label = paste0("$647k in capital invesment in the first\n", 
                                 "68  months needed to generate returns")) +
        scale_x_continuous(breaks = 2018:2026, name = "projection year") +
        scale_y_continuous(limits = c(0.5, 1.5)) +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Annual difference in cost between ", 
                       "baseline scenario and alternative scenario")) +
        theme_classic() +
        guides(size = FALSE) +
        theme(legend.position = "bottom", 
              axis.title.y = element_blank(), 
              axis.text.y = element_blank())
        
```

The color indicates which scenario is the lowest cost scenario for a particular 
year. The magnitudes of the difference between the two cashflows is represented 
by the size of the bubble; negative values imply alternative scenario is more 
expensive. It is also used as the label at the center of each bubble. Two 
important points are called out on the annotation of the plot: the breakeven 
month (68) and the month where the cumulative cashflow differences becomes 
positive (104), that is when the the cumulative cashflows of the baseline policy
exceed those of the alternative policy.


### Why I think this visualization is not effective

When I first saw this visualization, it was not clear to me what information it 
was trying to make it easier for me to understand. A good data visualization 
should reduce the mental load of the viewer to gain understanding, not increase 
it. I was eventually able to decipher what it was trying to communicate after 
looking at it for a while and reading the accompanying explanatory notes. 

Two more specific issues with the visualization are the use of bubbles and the 
choice of colors. The area of the bubbles is used to convey the relative sizes 
in the cashflow differences. However, it is harder for humans to reason about 
relative areas (2d) than relative lengths (1d). The visualziation unnecessarily 
projects one-dimensional data into two dimensions. For example, I know because 
of the value labels that the bubble for year 2025 is about 50% larger than the 
bubble for 2024, but I would not have guessed it from the size of the bubble.

The color used in the visualization conveys important information. It's the only
graphical element which tells the user the sign of the differences in the two
sets of cashflows. As such, choosing colors that help the viwer understand the 
direction of the cashflows would probably be better than choosing colors which 
make it look pretty.

## So what would be better

I've highlighted a few things I think are issues with the original visualization.
Next, I will try to address these issues to create what I think is a better 
visualization. I've seen examples in books by Tufte and others show how to 
improve a bad chart. Perhaps they can go straight from A to B, but I typically 
need to iterate. I thought it might be interesting or helpful to record the 
steps I went through.

### Iteration 1 - Bursting the Bubbles

So my first thought was to get rid of the bubble charts and use a column chart. 
I also wanted to give some context to the months which are highlighted in the 
call outs so I added a second x-axis. Important to note, this secondary x-axis 
has exactly the same scale as the primary one.

```{r}
ggplot(d, aes(py, delta, fill = scenario)) + 
  geom_bar(stat = "identity", width = 0.3) +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c("#8EE5EE", "#AB82FF")) +
  annotate("segment", x = 2017.5, xend = 2022.5, 
           y = 100, yend = 100, 
           color = "blue") +
  annotate("text", x = 2020, y = 175, 
           label = paste0("$647k in capital invesment in the first\n", 
                          "68  months needed to generate returns")) +
  annotate("text", x = 2024.5, y = -100, 
           label = paste0("TCOR breakeven at 68 momnths\n", 
                          "Realize gain after 104 months")) +
  scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = 12*(1:9))) +
  ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Annual difference in cost between ", 
                       "baseline scenario and alternative scenario")) +
  theme_classic() + 
  theme(legend.position = "none")+
  ylab("Annual Cashflow Difference (in thousands)") + xlab("projection year")

```


I think this already much better. It now immediately obvious that about half way
between 60 and 72 months the sign changes from a negative delta to a positive one.

### Iteration 2 - Clean up the Call Outs

Next, I thought it was sort of visually jarring to have the call out for the 
first half of the chart above the 0 line and the other below. I also thought it 
might be good to try to tie the call outs to the relate regions of the chart 
with color. 

```{r}
ggplot(d, aes(py, delta, fill = scenario)) + 
  geom_bar(stat = "identity", width = 0.3) +
  scale_fill_manual(values = c("#8EE5EE", "#AB82FF")) +
  geom_hline(yintercept = 0) +
  annotate("segment", x = 2017.5, xend = 2022.5, 
           y = 400, yend = 400, 
           color = "#AB82FF") +
  annotate("text", x = 2020, y = 455, 
           label = paste0("$647k in capital invesment in the first\n", 
                          "68  months needed to generate returns")) +
  annotate("segment", x = 2022.5, xend = 2026.5, 
           y = 400, yend = 400, 
           color = "#8EE5EE") +
  annotate("text", x = 2024.5, y = 455, 
           label = paste0("TCOR breakeven at 68 momnths\n", 
                          "Realize gain after 104 months") )+
  scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = c(12*(1:9), 68, 104))) +
  scale_y_continuous(limits = c(-260, 475)) +
  ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Annual difference in cost between ", 
                       "baseline scenario and alternative scenario")) +
  theme_classic() + 
  theme(legend.position = "none")+
  ylab("Annual Cashflow Difference (in thousands)") + xlab("projection year")
```

### Iteration 3 - Concentrate on the Cumulative Cashflows

There are two "critical points" that are called out on the chart. 68 months is 
the point at which the difference between the cashflows of the two scenarios 
flips from negative to postive; 104 months is the point past which all of the 
postive differences in cashflows exceed all negative ones, but that is not 
something that can be seen in this version of the visualization. This led me to 
overlay the cumulative cashflows.

```{r}
d_cumulative <- mutate(d, cumdelta = cumsum(delta), start = cumdelta - delta)

ggplot(d_cumulative, aes(x = py, fill = scenario)) + 
    geom_col(aes(y = delta), width = 0.3) + 
    scale_fill_manual(values = c("#8EE5EE", "#AB82FF")) +
    geom_point(aes(y = cumdelta)) + 
        geom_line(aes(y = cumdelta), color = "black") +
    scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future",
                                         breaks = sort(c(12*(1:9), 68, 104)))) +
    scale_y_continuous(breaks = c(-647, -500, -250, 0, 250, 500)) +
    geom_hline(yintercept = 0) +         
    theme_classic() + 
  ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Annual difference in cost between ", 
                       "baseline scenario and alternative scenario")) +
  theme_classic() + 
  theme(legend.position = "none")+
  ylab("Annual Cashflow Difference (in thousands)") + xlab("projection year")
```

Okay, now we're getting somewhere. The cumulative polot removes the need for 
any callouts since we can see clearly the two critical points in the 
cumulative data. The inflection point, the part where the curve bottoms out is 
the point where the cashflows change from negative to positive. And the place 
where the cumulative plot crosses the x-axis is visually near 104 months.

### Iteration 4 - What about a Waterfall?

I want to emphacise both the incremenetal changes and the cumulative values. 
My first attempt to accomplish this is to use a waterfall chart. 


```{r}
ggplot(d_cumulative) +
        geom_rect(aes(xmin = py - 0.2, 
                  xmax = py + 0.2, 
                  ymin = start, 
                  ymax = cumdelta,
                  fill = scenario)) +
        scale_fill_manual(values = c("#8EE5EE", "#AB82FF")) +
        geom_hline(yintercept = 0) +
        geom_hline(yintercept = -647) +
        scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = sort(c(12*(1:9), 68, 104)))) +
        scale_y_continuous(breaks = c(-647, -500, -250, 0, 250, 500)) +
        theme_classic() + 
        theme(legend.position = "none") +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Annual difference in cost between ", 
                       "baseline scenario and alternative scenario")) +
        ylab("Annual Cashflow Difference (in thousands)") + 
        xlab("projection year")
```

Now I can see how each incremental value accumulates to the cumulative value. The two 
"critical points" jump out of the plot. So there are three things now that bother me about 
this plot. 

- Although if I read the chart left to right, I can easily follow the incremental
movements, I find it a little confusing to understand the direction the waterfall is moving
if I start by looking at the middle of the plot.

- I wonder if there is a way to make what's happening here more grokable. The values being
plotted show the incremental and cumulative cost savings between the alternative scenario 
and the baseline scenario. Negative cost savings feels a little bit like a double negative.

- Now that the main feature of the plot calls out both the inflection point and the point 
at which the cumulative savings become positive, it's clear that the analysis was done 
using more granular data then what we're showing. If we could plot quarterly cashflows we 
could see the two critical points exactly.

I don't actually have the quarterly data, so I will try to address the first two points 
above using the annual data and then extrapolate the quarterly data.

### Iteration 5 - Changing Direction and Color

I'm going to see if I can deal with the "grokablness" issue first. The baseline 
scenario is what will happen if the client continues their current policy; the 
alternative involves a change to that policy. I want to frame things in terms of
costs and savings to the client relative to the baseline scenario. I'll start by
refactoring the data slightly.

```{r}
d_alt <- mutate(d_raw, delta = -delta) %>%
        rename(amt = delta) %>%
        mutate(direction = as.factor(case_when(amt >= 0 ~ "cost", TRUE ~ "savings")),
               cumamt = cumsum(amt),
               start = cumamt - amt)
gt(d_alt)
```


```{r}
ggplot(d_alt) +
        geom_rect(aes(xmin = py - 0.2, 
                  xmax = py + 0.2, 
                  ymin = start, 
                  ymax = cumamt,
                  fill = direction)) +
        scale_fill_manual(values = c("#FF6A6A", "#00CD66")) + 
        geom_hline(yintercept = 0) +
        scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = sort(c(12*(1:9), 68, 104)))) +
        scale_y_continuous(breaks = c(-500, -250, 0, 250, 500, 647)) +
        theme_classic() + 
        theme(legend.position = "right", legend.justification = "top")+
        ylab("Additional Cost/(Savings) (in thousands)") + 
        xlab("projection year") +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Additional Annual Cost/(Savings) from switching ", 
                       "from the baseline to alternative scenario")) 

```

So now, I think this gels with my intuition, the direction, labels, and color all point in 
the same direction: increasing expense is an increase and is showing in red (bad); 
decreasing expense is a decrease and is showing green (good). 


### Iteration 6 - Clear Directions

The choice of colors which indicate direction is helpful, but it doesn't completly address
my concern with being able to easily see the direction of each incremental change.

One idea is to make the width of each column wider and add data labels.

```{r}
ggplot(d_alt) +
        geom_rect(aes(xmin = py - 0.35, 
                  xmax = py + 0.35, 
                  ymin = start, 
                  ymax = cumamt,
                  fill = direction)) +
        scale_fill_manual(values = c("#FF6A6A", "#00CD66")) +
        geom_text(aes(x = py, y = cumamt - sign(amt) * 25, 
                      label = ifelse(abs(amt) < 50, "", amt))) +
        geom_hline(yintercept = 0) +
        scale_x_continuous(breaks = 2018:2026, expand = c(0, 0),
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = sort(c(12*(1:9), 68, 104)))) +
        scale_y_continuous(breaks = c(-500, -250, 0, 250, 500, 647)) +
        theme_classic() + 
        theme(legend.position = "right", legend.justification = "top") +
        ylab("Additional Cost/(Savings) (in thousands)") + 
        xlab("projection year") +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Additional Annual Cost/(Savings) from switching ", 
                       "from the baseline to alternative scenario")) 

```

An alternative would be to use something inheriantly directional like an arrow.

```{r}
ggplot(d_alt) +
        geom_segment(aes(x = py, 
                  xend = py, 
                  y = start, 
                  yend = cumamt,
                  color = direction), 
                  arrow = arrow(length = unit(0.25, "cm"), type = "closed")) +
        scale_color_manual(values = c("#FF6A6A", "#00CD66")) +
        geom_text(aes(x = py - 0.28, y = (start + cumamt)/2, 
                      label = ifelse(abs(amt) < 50, "", amt))) +
        geom_hline(yintercept = 0) +
        scale_x_continuous(breaks = 2018:2026,
                     sec.axis = sec_axis(~ 12*(. - 2017), 
                                         name = "Months into Future", 
                                         breaks = sort(c(12*(1:9), 68, 104)))) +
        scale_y_continuous(breaks = c(-500, -250, 0, 250, 500, 647)) +
        theme_classic() + 
        theme(legend.position = "none") +
        ylab("Additional Cost/(Savings) (in thousands)") + 
        xlab("projection year") +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Additional Annual Cost/(Savings) from switching ", 
                       "from the baseline to alternative scenario")) 

```

Okay, I like that even better than I thought I would. I don't think the legend 
is necessary with this version.

### Iteration 7 - Extrapolating the quarterly data

If we take a look at the annual cumulative data again, I think it's clear we can get a pretty
good fit with a quadratic or cubic fit.

```{r}

d_model <- select(d_alt, py, cumamt) %>%
        mutate(py = py + 1) %>%
        bind_rows(tibble(py = 2018, cumamt = 0)) %>%
        arrange(py)

model <- lm( cumamt ~ poly(py, 3), data = d_model)

summary(model)

predicted_data <- data.frame(py = 2018:2027) %>%
        mutate(amt = predict(model, .))

ggplot(d_model, aes(x = py, y = cumamt)) + 
        geom_point() +
        geom_line(aes(x = py, y = amt), data = predicted_data, color = "blue")
```


So I'll now use this model to create simulated quarterly data.

```{r}
years <- 2018:2026
num_years <- length(years)
d_qtr <- data.frame(py = c(rep(years, 4), 2027), 
                    qtr = c(rep(0, num_years), 
                            rep(0.25, num_years), 
                            rep(0.5, num_years), 
                            rep(0.75, num_years), 0)) %>%
        mutate(py = py + qtr) %>%
        select(py) %>% 
        arrange(py) %>%
        mutate(cumamt = predict(model, .), month = 4 * row_number())

d_qtr$cumamt[1] <- 0
d_qtr <- mutate(d_qtr, 
                amt = cumamt - lag(cumamt),
                direction = as.factor(case_when(amt >= 0 ~ "cost", 
                                                TRUE ~ "savings")),
                start = cumamt - amt)

gt(d_qtr)
```


```{r}
ggplot(d_qtr) +
        geom_segment(aes(x = py, 
                  xend = py, 
                  y = start, 
                  yend = cumamt,
                  color = direction), 
                  arrow = arrow(length = unit(0.15, "cm"), type = "closed")) +
        scale_color_manual(values = c("#FF6A6A", "#00CD66")) +
        geom_hline(yintercept = 0) +
        scale_x_continuous(breaks = 2018:2027,
                     sec.axis = sec_axis(~ 12*(. - 2017),
                                         name = "Months into Future", 
                                         breaks = sort(c(12*(1:10), 68, 104)))) +
        scale_y_continuous(breaks = c(-500, -250, 0, 250, 500, 647)) +
        theme_classic() + 
        theme(legend.position = "none") +
        ylab("Additional Cost/(Savings) (in thousands)") + 
        xlab("projection year") +
        ggtitle("Projection of Future Cost: Baseline vs Alternative", 
                paste0("Additional Quarterly Cost/(Savings) from switching ", 
                       "from the baseline to alternative scenario")) 

```


Although the quarterly data doesn't line up exactly with the critical points, this
is due to approximating the data.

## Final Verdict

After going through this process, I think the either of the alternatives from
iteration 6 would be good choices for the final version of this visualization. I
think the quarterly data version is inferior for two reasons. First, now that 
there are so many points it really looks like a line graph, similiar to what I
created to show the polynomial fit and it feels "crowded". Secondly, as I have 
thought more about the cashflows we're describting in this visualization, and,
more importantly, the consumers of the visualization, I think they might more
natuarally think of these cashflows as annual.
